---
title: "A1-Basic"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Task 1.A – Exploration
## Data preprocessing
We kept the following features in our preprocessed dataset:
```{r}
data = read.csv('ODI-2018_clean.csv')
colnames(data)
```
The study program feature was cleaned using the 'format_study_program.R' script. The course features were relatively clean already and did not need additional cleaning. This was also the case for the 'gender', 'chocolate makes you' and 'give a random number' features. The original 'birthday' feature was cleaned using the 'clean_birthday_bedtime.py' script and splitted into a day, month and year feature. Unfortunately, the 'bedtime' feature cleaning was problematic and was manually formatted. Finally, the 'good day' features were cleaned using the 'GoodDay_cleanup.R' script.

## Exploratory data analysis
The cleaned data consisted of 218 samples and 14 features:
```{r}
nrow(data)
ncol(data)
```

More men participate in the data mining techniques course:
```{r}
pie(table(data$What.is.your.gender.), col = rainbow(3), labels = c("Unknown: 4","Females: 63","Males: 150"))
```
\
Figure 1. Pie chart showing the gender distribution in the data mining techniques course.\
\

The birth years distribution (left out two students who were born in 1768 and 1931):
```{r}
plot(table(data$birth_year), xlab = "Year", ylab = "Number of students", xlim = c(1981,2000))
```
\

Figure 2. Birth year distribution in the data mining techniques course.\
\

Interesting information about chocolate:
```{r}
choc = table(data$Chocolate.makes.you.....)
names(choc) = c("NA", "fat", "No idea", "Neither", "slim", "unknown")
barplot(choc)
```
\

Figure 3. Chocolate makes you.... Answer frequencies.

## Basic classification/regression
Let's try to predict the gender of samples using the features in the datases.

```{r}
library(caret)
library(doMC)
# parallel processing: set number of cores
registerDoMC(cores = 4)
# split for train/test data
trainIndex = createDataPartition(data$Survived, p = .7, list = FALSE)
train = data[ trainIndex,]
test  = data[-trainIndex,]
```